Cross Validation Approaches for Penalized Cox Regression

### 2019-04-03 main.tex
edit 3.1

\begin{figure}[ht]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_2_new.png}
  \caption{\label{Fig:mse-brier-c} Simulation comparing 4 CV methods. The horizontal axis in all three plots is $s$, which represents the signal strength in the data. The value of $s$ is varied from 0.4 to 0.9.  $\log(\text{MSE ratio})$, relative to the oracle modle is plotted in the left panel. Out-of-sample Brier scores are plotted in the middle. Out-of-sample Harrell's C index is plotted in the right panel. For each simulated data set, n = 120, p = 1000. Expected censoring percentage is 10$\%$. \note{I really think the x axis should be Signal (s) or Signal strength (s)}}
\end{figure}

@@

\subsection {Stability}
\label{Sec:stability}
  \bnote{still working on a new simulation}
\par In this section, we present a small simulation study to illustrate the numerical stability issues faced by standard cross-validation in the partial likelihood setting.  The standard approach is both straightforward and enjoys very good predictive accuracy, as seen in Section~\ref{Sec:accuracy}. The main drawback to its widespread use is that the number of uncensored observations in the test set may be insufficient to construct a well-defined partial likelihood, leaving the overall estimate of cross-validation error undefined as well.

To illustrate how common this problem is, and what factors it depends on, we conducted a simulation study varying the number of folds and censoring rate. Throughout we fixed the number of observations at $n = 100$. When varying the censoring rate, we fixed the number of folds at $K = 10$; when varying the number of folds, we fixed the censoring rate at 10\%. We generated 200 independent data sets according to the mechanism described in Section~\ref{Sec:accuracy} and recorded whether or not standard cross-validation was able to be carried out (i.e., whether the partial likelihood was well-defined for all folds).

\note{did we make an effort to balance censoring across folds here?  we should comment on this}

\note{also, hmm, censoring rate isn't actually fixed at, say, 30\%...eh, maybe it's OK}

\par As illustrated in Figure~\ref{Fig:stability}, at $n=100$ the standard approach can be problematic even at fairly modest censoring rates.  For example, standard CV cannot be used with over 25\% of data sets when half of all observations are censored.  Furthermore, even if the censoring rate is just 10\%, standard CV strongly limits the number of folds that can be included in the cross-validation process, becoming very unstable when the number of folds exceeds 10.

\begin{figure}[h]
  \centering
  \includegraphics[height= 7cm ]{./figures/figure_3_new.png}
  \caption{\label{Fig:stability}Comparing Stability \note{This plot needs better y axis labels and caption; also, these are not percents.}}
\end{figure}

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_2_new.png}
  \caption{\label{Fig:mse-brier-c} Simulation comparing 4 CV methods. The horizontal axis in all three plots is $s$, which represents the signal strength in the data. The value of $s$ is varied from 0.4 to 0.9.  $\log(\text{MSE ratio})$, relative to the oracle modle is plotted in the left panel. Out-of-sample Brier scores are plotted in the middle. Out-of-sample Harrell's C index is plotted in the right panel. For each simulated data set, n = 120, p = 1000. Expected censoring percentage is 10$\%$. 
%\note{I really think the x axis should be Signal (s) or Signal strength (s)}
}
\end{figure}

@@

\subsection {Stability}
\label{Sec:stability}
 \bnote{I updated the writing and figure~\ref{Fig:stability} for this section}

The standard cross-validation in the partial likelihood setting is both straightforward and enjoys very good predictive accuracy, as seen in Section~\ref{Sec:accuracy}. However there are two main stability drawbacks to its widespread use. The following simulation studies illustrated those issues faced by the standard approach.

The first stability issue with the standard approach is that the number of uncensored observations in the test set may be insufficient to construct a well-defined partial likelihood, leaving the overall estimate of cross-validation error undefined as well. When computing the partial likelihood, only individuals whose events are observed will contribute to the likelihood. If there is no event observed in a particular fold, then the partial likelihood for that fold is undefined. To illustrate how common this problem is, we conducted a simulation study. We generated 1000 independent data sets of $n = 100$ observations with fixed censoring rate. For each data set, we conducted a 10-fold cross-validation and recorded whether or not there was at least one event in each fold (i.e., whether the partial likelihood was well-defined for all folds). We repeated the same mechanism for scenarios with different censoring rates, ranging from $40\%$ to $80\%$. As illustrated in the left panel in Figure~\ref{Fig:stability}, at $n=100$ the standard approach can be problematic even at modest censoring rates. For example, standard CV cannot be used with over 20\% of data sets when $70\%$ of all observations are censored.

One way could be used to get around the undefined partial likelihood for the standard approach is to balance the censoring status across all folds when making fold assignments. That is to make sure the number of events observed is about the same across all folds. Unless the number of the folds exceeds the number of observed events, such as the leave-one-out cross-validation, this approach will guarantee the partial likelihood in each fold is well-defined for the standard approach, thus the overall cross-validated error is well-defined.

Even if the partial likelihood could be well-defined with balanced fold assignment, there is another drawback of the standard approach: it tends to have more uncertainty in its selection. The predictive performance of the model selected by this approach has greaterr variation than the other approaches, especially when there are only a few events available to construct the partial likelihood.

We designed the following simulation study to illustrate this issue. We generated data sets with $n = 120$ observations. The censoring rate is fixed exactly at $50\%$, thus there are 60 events observed among all observations. For each data set, we conducted the balanced fold assiggment with number of folds set to be 10, 12, 15, 20, 30 and 60. When the number of folds was set to be 10, there were exactly 6 events in each fold; when the number of folds was set to be 60, there were one event in each fold. After each fold assignment, we computed the cross-validated error using all four approaches described in Section~\ref{Sec:accuracy}. We recorded the model that minimizes the cross-validated error for each approach and computed its squared error loss. We replicated the whole process for $N = 200$ time. In the end, instead of looking at the mean as in Section~\ref{Sec:accuracy}, we computed the standard deviation of the squared error loss for each approach.

The results of this simulation are illustrated in the right panel in Figure~\ref{Fig:stability}. While all other three approaches are not affected by the number of folds and number of events in each fold, the variability of the model performance has drastically changed for the standard approach. When there are sufficient number of events in each fold, the standard approach perform pretty close to the other three approaches. However, when there is only one or two events in each fold, the standard approach has much more variability in its predictive performance. Even if we can still use the standard approach when there are few number of events, there is a bigger uncertainty in terms of how well the predictive performance would be.

Based on the two simulations illustrated in this section, we would make the following recommendations: i) always balance the censoring status when conducting fold assignments for the standard approach, ii) avoid using the standard approach in conjuction with high number of folds, iii) avoid using the standard approach when there are few number of events observed.

\begin{figure}[h]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_3.png}
  \caption{\label{Fig:stability} Simulations illustrating the stability issues with the standard approach. The figure on the left is generated with $n = 100$ observations. The horizontal axis is the censoring rate. The vertical axis is the proportion of undefined partial likelihood when the fold assignment is not balanced. The figure on the right is generated with $n = 120$. The fold assignment is balanced in this simulation so that the number of events in each fold is the same. The horizontal axis is the number of events per fold. The vertical axis is the standard deviation of the squared error loss. All four methods were compared in this simulation. The standard approach would have much greater variability when there are only few events in the fold.}
\end{figure}
